---
title: 'Reproducible Research: Peer Assessment 1'
author: "Luiz Carlos Moutinho Pataca"
date: "December 14, 2016"
output:
    html_document:
        keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Cleaning environment
rm(list = ls())

```

## Loading and preprocessing the data


```{r load_dada}
#Loading compressed file "activity.zip" with read.csv() function, and defining
#class for each variable.
activity <- read.csv(file = unzip(zipfile = "activity.zip", files = "activity.csv"),
                     header = TRUE, na.strings = "NA",
                     colClasses = c("integer", "Date", "integer"))
```

## What is mean total number of steps taken per day?

```{r total_per_day}
# Calculating total number of steps taken per day. Missing values are ignored.
total_day <- aggregate(activity$steps, by = list(activity$date), sum)
names(total_day) <- c("date", "steps")
```

```{r histogram_total}
# Creating a histogram of total number of steps taken each day.
hist(total_day$steps,
     main = "Histogram of Total Number of Steps Taken Each Day",
     xlab = "Steps")
```


```{r Mean_Median}
# Calculating and reporting the mean and median of the total number of steps
# taken per day.
Mean <- mean(total_day$steps, na.rm = TRUE)
Median <- median(total_day$steps, na.rm = TRUE)

library(knitr)
kable(cbind(Mean, Median),
      caption = "Mean and Median of the total number of steps taken per day",
      align = c("l", "l"))
```


## What is the average daily activity pattern?

Using  function aggregate() for calculating 5-minute interval mean across all days.

```{r Daily_Mean}
daily_mean <- aggregate(x = activity$steps, by = list(activity$interval),
                        FUN = "mean", na.rm = TRUE)
names(daily_mean) <- c("interval", "steps")

# Ploting daily average
plot(x = daily_mean$interval, y = daily_mean$steps, type = "l",
     xlab = "5-minute interval", ylab = "Steps taken average",
     main = "Daily Average Activity Pattern")
```

Determining the 5-minute interval with the largest number of steps.

```{r interval_max}
id_max <- which.max(daily_mean$steps)
Interval_max <- daily_mean$interval[id_max]
```

***5-minute Interval with maximum number of steps: `r Interval_max`***

## Imputing missing values

Calculating and reporting the total number of missing values in the dataset.

```{r missing_values}
missing_values <- sum(is.na(activity$steps))
```

***Total missing values in the dataset: `r missing_values`***

Using 5-minute interval total average for filling missing values in the dataset.

```{r activity_fill}
activity_fill <- merge(x = activity, y = daily_mean, by="interval", suffixes=c("",".mean"))
id_na <- is.na(activity_fill$steps)
activity_fill$steps[id_na] <- activity_fill$steps.mean[id_na]
activity_fill <- activity_fill[order(activity_fill$date, activity_fill$interval), c(2, 3, 1)]
activity_fill$steps <- as.integer(round(x = activity_fill$steps, digits = 0))
```

**Properties and Summary of original dataset**

```{r summary_ori}
str(activity)
summary(activity)
```

**Properties and Summary of dataset with filled missing values**

```{r summary_fill}
str(activity_fill)
summary(activity_fill)
```

Calculating total number of steps taken per day. Missing values are filled in.

```{r total_per_day_fill}
total_day_fill <- aggregate(activity_fill$steps, by = list(activity_fill$date), sum)
names(total_day_fill) <- c("date", "steps")
```

Creating a histogram of total number of steps taken each day. Missing values are filled in.

```{r histogram_total_fill}
hist(total_day_fill$steps,
     main = "Histogram of Total Number of Steps Taken Each Day",
     xlab = "Steps")
```

Calculating and reporting the mean and median of the total number of steps taken per day. Missing values are filled in.

```{r Mean_Median_fill}
Mean <- mean(total_day_fill$steps, na.rm = TRUE)
Median <- median(total_day_fill$steps, na.rm = TRUE)

library(knitr)
kable(cbind(Mean, Median),
      caption = "Mean and Median of the total number of steps taken per day. Missing values are filled in.",
      align = c("l", "l"))
```


The histogram and calculated values of Mean and Median show that the impact of imputing missing data on the estimates of the total daily number of steps is rather low.

## Are there differences in activity patterns between weekdays and weekends?


```{r total_weekday}
# First defining english locale
Sys.setlocale(category = "LC_ALL", locale = "en_US.UTF-8")
# Creating a new factor variable indicating whether a given date is a weekday or
# weekend day
weekday <- weekdays(activity_fill$date)
id_weekend <- which(weekday == "Saturday" | weekday == "Sunday")
weekday[id_weekend] <- "weekend"
weekday[-id_weekend] <- "weekday"
activity_fill$weekday <- as.factor(weekday)
# Calculating total average of number of steps for weekdays and weekends
total_weekday <- aggregate(steps ~ interval + weekday, data = activity_fill, mean)
# Plotting
library("lattice")
xyplot(steps ~ interval | weekday, total_weekday, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Average of steps")
```

The plots show differences between weekdays and weekends. On weekdays more steps are taken before 10:00 am compared to weekends.
